HO unification from object language to meta language

### 2024-02-22 paper/main.tex
update tex

  The object logic we are interested in is Coq's~\cite{Coq-refman}
  Dependent Type Theory (DTT),
  for which we aim to implement a unification procedure \Uo using the ML
  Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
  Elpi's equational theory comprises
  $\eta\beta$ equivalence and comes equipped with a
  higher order unification procedure \Ue restricted to the pattern
  fragment~\cite{miller92jsc}.
  We want \Uo to be as powerful as \Ue but on the object logic DTT.
  Elpi also comes with an encoding for DTT that works well
  for meta-programming~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
  Unfortunately this encoding, which we refer to as \Fo,
  ``underuses'' \Ue by restricting it to first-order unification problems only. 
  To address this issue, we propose a better-behaved encoding, \Ho,
  demonstrate how to map unification problems in \Fo
  to related problems in \Ho, and illustrate
  how to map back the unifiers found by \Ue, effectively implementing
   \Uo on top of \Ue for the encoding \Fo.

@@

\noindent
Unfortunately this direct translation of rule (r3) uses the
predicate \coqIn{P} essentially as a first order term for the meta
language (its type is \elpiIn{tm}).
If we try to backchain the rule (r3) on the encoding of the goal (g) above:

@@

\paragraph{Contributions}
In this paper we discuss alternative encodings of Coq in
Elpi~\ref{sec:encodings}, then we identify a minimal language \Fo
in which the problems sketched here can be fully described.
We then detail an encoding \elpiIn{comp} from \Fo to \Ho (the language of
the meta language) and a decoding \elpiIn{decomp} to relate the unifiers
bla bla.. TODO citare Teyjus.

@@

To state precisely the problem we solve we need a \Fo representation
of DTT terms and a \Ho one.
We call \Eo the equality over ground terms in \Fo,
\Ee the equality over ground terms in \Ho,
\Uo the unification procedure we want to implement and
\Ue the one provided by the meta language.
TODO extend \Eo and \Ee with reflexivity on uvars.

@@

We illustrate a compilation $\C{s}{t}{l}$ that
maps a term $s$ in \Fo to a term $t$ in \Ho and a list of links $l$.
The links connect unification variables in \Ho with variables
in \Fo and are used to decompile the assignment,
$\D{\sigma}{l}{\rho}$.

@@

@@

\begin{coqcode}
  forall (t  u : term_ol) (s : subst_ol) 
         (t' u': term_ml) (s': subst_ml), 
    equal_ol t u s <->
      comp t t' /\ comp u u' /\ 
      unif_ml t' u' s' /\ decomp s' s.
\end{coqcode}

@@

The compilation step is meant to recover the higher-order variables of the OL,
expressed in a first order way, by replacing them with higher-order variables in
the ML. In particular, every time a variable of the OL is encountered in the
original term, it is replaced with a meta variable, and if the OL variable is
applied to a list $L$ of terms, then this list becomes the scope of the variable
if $L$ is a list of distinct names. For all the other constructors of
\elpiIn{tm}, the same term constructor is returned and its arguments are
recursively compiled. The predicate in charge for term compilation is:\\
\elpiIn{type comp tm -> tm -> links -> links -> subst -> subst -> o}.
Where, we take the term of the OL, produce the term of the ML, take a list
of link and produce a list of new links, take a substitution and return a 
new substitution.

@@

The following code
\begin{elpicode}
  kind link type.
  type link nat -> nat -> nat -> subst.
\end{elpicode}

\noindent
defines a link, which is a relation between to variables indexes, the first
being the index of a OL variable and the second being the index of a ML
variable. The third integer\todo{integer or nat?} is the number of term in the
scope of the two variables, or equivalently, in a typed language, their arity.

>>>>

  The object logic we are interested in is Coq's~\cite{Coq-refman}
  Dependent Type Theory (DTT),
  for which we aim to implement a unification procedure \Uo using the ML
  Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
  Elpi's equational theory comprises
  $\eta\beta$ equivalence and comes equipped with a
  higher order unification procedure \Ue restricted to the pattern
  fragment~\cite{miller92jsc}.
  We want \Uo to be as powerful as \Ue but on the object logic DTT.
  Elpi also comes with an encoding for DTT that works well
  for meta-programming~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
  Unfortunately this encoding, which we refer to as \Fo,
  ``underuses'' \Ue by restricting it to first-order unification problems only. 
  To address this issue, we propose a better-behaved encoding, \Ho,
  demonstrate how to map unification problems in \Fo{}
  to related problems in \Ho, and illustrate
  how to map back the unifiers found by \Ue, effectively implementing
   \Uo on top of \Ue for the encoding \Fo.

@@

\noindent
Unfortunately this direct translation of rule (r3) uses the
predicate \coqIn{P} essentially as a first order term for the meta
language (its type is \elpiIn{tm}).
If we try to backchain the rule (r3) on the encoding of the goal (g) below:

@@

\paragraph{Contributions}
In this paper we discuss alternative encodings of Coq in
Elpi~\ref{sec:encodings}, then we identify a minimal language \Fo
in which the problems sketched here can be fully described.
We then detail an encoding \elpiIn{comp} from \Fo{} to \Ho (the language of
the meta language) and a decoding \elpiIn{decomp} to relate the unifiers
bla bla.. TODO citare Teyjus.

@@

To state precisely the problem we solve we need a \Fo{} representation
of DTT terms and a \Ho one.
We call \Eo the equality over ground terms in \Fo,
\Ee the equality over ground terms in \Ho,
\Uo the unification procedure we want to implement and
\Ue the one provided by the meta language.
TODO extend \Eo and \Ee with reflexivity on uvars.

@@

We illustrate a compilation $\C{s}{t}{l}$ that
maps a term $s$ in \Fo{} to a term $t$ in \Ho and a list of links $l$.
The links connect unification variables in \Ho with variables
in \Fo{} and are used to decompile the assignment,
$\D{\sigma}{l}{\rho}$.

@@

\

@@

\begin{coqcode}
  forall (t  u : term_ol) (s : subst_ol) 
         (t' u': term_ml) (s': subst_ml), 
    equal_ol t u s <->
      comp t t' /\ comp u u' /\ 
      unif_ml t' u' s' /\ links ... /\ decomp s' s.
\end{coqcode}

@@

The compilation step is meant to recover the higher-order variables of the OL,
expressed in a first order way, by replacing them with higher-order variables in
the ML. In particular, every time a variable of the OL is encountered in the
original term, it is replaced with a meta variable, and if the OL variable is
applied to a list of distinct names $L$, then this list becomes the scope of the variable. 
For all the other constructors of
\elpiIn{tm}, the same term constructor is returned and its arguments are
recursively compiled. The predicate in charge for term compilation is:\\
\elpiIn{type comp tm -> tm -> links -> links -> subst -> subst -> o}.
Where, we take the term of the OL, produce the term of the ML, take a list
of link and produce a list of new links, take a substitution and return a 
new substitution.

@@

The following code
%
\begin{elpicode}
  kind link type.
  type link nat -> nat -> nat -> subst.
\end{elpicode}
%
\noindent
defines a link, which is a relation between to variables indexes, the first
being the index of a OL variable and the second being the index of a ML
variable. The third integer\todo{integer or nat?} is the number of term in the
scope of the two variables, or equivalently, in a typed language, their arity.

As an example, let's study the following unification problem (a slightly 
modified version from \cref{sec:intro}):

\begin{elpicode}
  lam x\ app[c"decision", app[c"nfact", x, c"3"]] ~\Uo~
    lam x\ app [c"decision", app[uv 0, x]]
\end{elpicode}

\noindent
we have the main unification problem where the nested \elpiIn{app} nodes have
lists of different lengths making the unification to fail. The compilation of 
these terms produces a new unification problem with the following shape:

\begin{elpicode}
  lam x\ app[c"decision", app[c"nfact", x, c"3"]] ~\Ue~ 
    lam x\ app [c"decision", uv 1 [x]]
\end{elpicode}

\noindent
The main difference is the replacement of the subterm \elpiIn{app[uv 0, x]} of
the OL with the subterm \elpiIn{uv 0 [x]}. Variable indexes are chosen by the
ML, that is, the index \elpiIn{0} for that unification variable of the OL term
has not the sam meaning of the index \elpiIn{0} in the ML. There exists two
different substitution mapping, one for the OL and one for the ML and the indexes
of variable point to the respective substitution.
